---
title: "DGE_analysis_sem1"
author: "jayden beckwith"
date: "2023-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries 

```{r setup, include=TRUE}
#libraries
library(dplyr)
library(tidyverse)
library(DESeq2)
library(biomaRt)
library(tibble)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(ggplot2)
library(data.table)
library(ggrepel)
library(reticulate)
library(VennDiagram)
library(plotly)
library(limma)
library(ggpubr)
library(edgeR)
library(gridExtra)
library(statmod)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(OUTRIDER)
```

### rep dfs

creating dfs of technical replicates - 8 technical reps - 3 batches - muscle/blood

```{r}
test_meta <- fread("test_metadata_DGE_checkpoint.csv")
meta_df <- test_meta[,1:7]
meta_df
```
```{r}
test_count_mat <- fread('final_count_mat.csv')
test_counts <- data.frame(test_count_mat)
test_counts

#check for any dups - if any get index
anyDuplicated(test_counts$Geneid)
test_counts$Geneid[which(duplicated(test_counts$Geneid))]
```
```{r}
#drop dup and impute null counts as 0
test_dup <- test_counts[-14998, ]
test_dups <- data.frame(test_dup, row.names = 1)
test_dups[is.na(test_dups)] <- 0
test_dups
```

###  Collapse replciates
```{r}
#factor cols and relevel
meta_df$BATCH<- factor(meta_df$BATCH)

meta_df$COLLECTION_TYPE <- factor(meta_df$COLLECTION_TYPE)

meta_df$MND <- factor(meta_df$MND)


meta_df$SEX <- factor(meta_df$SEX)
meta_df
```
```{r}
#create deseq2 matrix and design 
dds <- DESeqDataSetFromMatrix(countData = test_dups,
                              colData = meta_df,
                              #account for batch effect 
                              design = ~ BATCH + MND)

#collapse technical replicates
dds_cols <- collapseReplicates(dds, dds$ID, dds$RUN)
dds_cols
```

### Data preprocessing and exploratory analyses 

```{r}
#filter lowly expressed genes 
keep <- rowSums(counts(dds_cols)) >= 10
dds_cols <- dds_cols[keep,]
dds_cols

#re-level and set MND status ref as control 
dds_cols$MND<- relevel(dds_cols$MND, ref = "Control")
dds_cols$MND
```
## PCA / scaling / normalisation / log transformation 

```{r}
#perform scaling 
est_dds <- estimateSizeFactors(dds_cols)

#normalise data and summarise cols for PCA
se <- SummarizedExperiment(log2(counts(est_dds, normalized=TRUE) + 1),
                           colData=colData(est_dds))
#plot PCA 
pca_d1 <- plotPCA(DESeqTransform(se), intgroup=c("COLLECTION_TYPE", "SEX", "MND"), returnData=TRUE)

percentVar <- round(100 * attr(pca_d1, "percentVar"))
pca_plot <- ggplot(pca_d1, aes(PC1, PC2, color=MND, shape=SEX, group=COLLECTION_TYPE)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of PAX Blood vs Muscle Biopsy Samples") + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_ellipse(linetype=2, aes(PC1, PC2, color=COLLECTION_TYPE)) + stat_ellipse(linetype=2, aes(PC1, PC2, color=COLLECTION_TYPE)) 
```



```{r}
pca_plot
```

### check for batch effects - VST

```{r}
vst <- vst(est_dds)

plotPCA(vst, intgroup=c("BATCH", "MND"))

assay(vst) <- limma::removeBatchEffect(assay(vst), vst$BATCH)

plotPCA(vst, intgroup=c("BATCH", "MND"))
```
##Filt Blood & Muscle

```{r}
filt_b <- est_dds[,est_dds$COLLECTION_TYPE == "Pax"]

#filter muscle
filt_mu <- est_dds[,est_dds$COLLECTION_TYPE == "Muscle"]
filt_mu
```
##PCA Blood and Muscle No Batch Effect
```{r}
vst_blood <- vst(filt_b)

plotPCA(vst_blood, intgroup=c( "MND", "SEX"))

assay(vst_blood) <- limma::removeBatchEffect(assay(vst_blood), vst_blood$BATCH)

blood_pca <- plotPCA(vst_blood, intgroup=c( "MND", "SEX"), returnData = T)

percentVar <- round(100 * attr(blood_pca, "percentVar"))
pca_blood <- ggplot(blood_pca, aes(PC1, PC2, color=MND, shape=SEX)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of Blood") + 
  theme(plot.title = element_text(hjust = 0.5)) #+ 
  #stat_ellipse(
   # data = blood_pca, aes(fill = factor(MND)),
   # geom = "polygon", alpha = 0.1, level = .95, type = "t", linetype = 2, show.legend = F
 # ) 

##PCA muscle 
vst_muscle <- vst(filt_mu)

plotPCA(vst_muscle, intgroup=c( "MND", "SEX"))

assay(vst_muscle) <- limma::removeBatchEffect(assay(vst_muscle), vst_muscle$BATCH)

muscle_pca <- plotPCA(vst_muscle, intgroup=c( "MND", "SEX"), returnData = T)

percentVar <- round(100 * attr(muscle_pca, "percentVar"))
pca_Muscle <- ggplot(muscle_pca, aes(PC1, PC2, color=MND, shape=SEX)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of Muscle") + 
  theme(plot.title = element_text(hjust = 0.5)) 
#+ 
 # stat_ellipse(
    #data = muscle_pca, aes(fill = factor(MND)),
    #geom = "polygon", alpha = 0.1, level = .95, type = "t", linetype = 2, show.legend = F
 # ) 

#join plot
ggarrange(pca_blood, pca_Muscle, ncol =2, align = "h", common.legend = TRUE, legend="bottom")
```

##DGE Analysis - Whole Dataset


```{r}
deseqdds <- DESeq(est_dds)
``` 

```{r}
dseq_res <- DESeq2::results(deseqdds)
```

```{r}
summary(dseq_res)
```
```{r}
resultsNames(deseqdds)
```
```{r}
res.05 <- DESeq2::results(deseqdds)
table(res.05$padj < 0.05)
```
```{r}
#set dseq results as df
dseq_df <- as.data.frame(dseq_res)

head(dseq_df)
```

```{r}
#use tibble to set rowname to col for edit
df <- tibble::rownames_to_column(dseq_df, "ID")
#need to be trimmed using regex and tidy for ensembl ids 
count_trim <- df %>% mutate(ensembl_gene_id = gsub("\\..*","", ID))
nrow(count_trim)
head(count_trim)

#BIomart db - get annotations 
listMarts()

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
listDatasets(ensembl) %>% filter(grepl("Human",description))

ensembl.con <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", host="https://useast.ensembl.org")
attr <- listAttributes(ensembl.con)
filters <- listFilters(ensembl.con)

#id, gene type, gene name
gene_df <- as.data.frame(getBM(attributes = c('ensembl_gene_id',"gene_biotype", 'external_gene_name'),
      filters="ensembl_gene_id",
      values = count_trim$ensembl_gene_id,
      mart = ensembl.con))

#merge dfs via ensemble ids 
ensembl_DEGs <- merge(count_trim, gene_df, by='ensembl_gene_id', all='TRUE')
#filt nans
filt_ensembl <- drop_na(ensembl_DEGs)
filt_ensembl <- data.frame(filt_ensembl, row.names = 2)
#filt_ensembl <- filt_ensembl %>% distinct(external_gene_name, .keep_all = TRUE)
head(filt_ensembl)

```
## Upregulated and downregulated genes

```{r}
#upregulated genes at LFC > 1 threshold and sig by p adj
upreg <- filt_ensembl %>% filter(log2FoldChange > 1 & padj < 0.05)
#no upreg genes
nrow(upreg)

downreg <- filt_ensembl %>% filter(log2FoldChange < -1 & padj < 0.05)
nrow(downreg)

```
```{r}
#dispersion plot
plotDispEsts(deseqdds, main='Dispersion Plot')
```
```{r}
DESeq2::plotMA(dseq_res, ylim=c(-2,2))
```

##Whole data - Top 50 DEGs

```{r}
#subset to get only protein coding genes
pc_genes <- subset(filt_ensembl, gene_biotype=='protein_coding')

#get top 50 protein coding DEGs
top_50_genes <- head(pc_genes[order(pc_genes$padj),], 50)
#set as df
top_50_genes <- as.data.frame(top_50_genes)
top_50_genes
```
##Volcano Plot of PCGs

```{r Volcano1, fig.height = 8, fig.width = 6.5, fig.align = "center"}
lab_italics <- paste0("italic('", lab = pc_genes$external_gene_name, "')")
  selectLab_italics = paste0(
    "italic('",
    c('COL19A1', "SAA1","SAA2","CDKN1A", "FGFBP2"),
    "')")

d1 <- EnhancedVolcano(pc_genes,
    lab = lab_italics,
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = selectLab_italics,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 1e-4,
    title = 'Volcano Plot',
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 6.0,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    parseLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'bottom',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')


d1
```

```{r}
#normalise counts for matrix 
mat <- counts(deseqdds, normalized = T)[rownames(top_50_genes),]
#get zscores for new matrix scale by 1 when transposing 
mat_z <- t(apply(mat,1,scale))
#set col names for new df
colnames(mat_z) <-rownames(dds_cols@metadata)
rownames(mat_z) <- top_50_genes$external_gene_name
```


```{r}

ann <- data.frame(SEX = dds_cols$SEX,
                  ALS= dds_cols$MND, 
                  COLLECTION_TYPE = dds_cols$COLLECTION_TYPE, stringsAsFactors = F)

colours <- list(
    SEX = c('Male' = 'blue', 'Female' = 'green'),
    ALS = c("Case" = "yellow","Control" = "orange"),
    COLLECTION_TYPE = c("Pax" = "purple", "Muscle" = "pink"))
colAnn <- HeatmapAnnotation(
    df = ann,
    which = "col", # 'col' (samples) or 'row' (gene) annotation?
    na_col = 'white', # default colour for any NA values in the annotation data-frame, 'ann'
    col = colours,
    annotation_height = 0.6,
    annotation_width = unit(1, 'cm'),
    gap = unit(1, 'mm'),
    annotation_legend_param = list(
      SEX = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'Sex',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface='bold'),
        labels_gp = gpar(fontsize = 8)),
      ALS = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'ALS',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface = "bold"),
        labels_gp = gpar(fontsize = 8)),
      COLLECTION_TYPE = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'Tissue',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface = "bold"),
        labels_gp = gpar(fontsize = 8))
      ))

#make heatmap for clustering 
DGE_heatmap <- Heatmap(mat_z, 
                       cluster_rows = T, 
                       cluster_columns = T, 
                       column_labels = colnames(mat_z),
                       name = "z-score",
                       column_title = "Samples",
                       row_title = "Genes", 
                       show_row_names = T,
                       show_column_names = F,
                       top_annotation = colAnn)

DGE_heatmap
```
##Blood DGE analysis

```{r}
blood_deseq <- DESeq(filt_b)
```



