---
title: "DGE_analysis_sem1"
author: "jayden beckwith"
date: "2023-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries 

```{r setup, include=TRUE}
#libraries
library(dplyr)
library(tidyverse)
library(DESeq2)
library(biomaRt)
library(tibble)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(ggplot2)
library(data.table)
library(ggrepel)
library(reticulate)
library(VennDiagram)
library(plotly)
library(limma)
library(rstatix)
library(ggpubr)
library(edgeR)
library(gridExtra)
library(statmod)
library(affy)
library(scales)
library(clusterProfiler)
library(org.Hs.eg.db)
library(variancePartition)
library(AnnotationDbi)
library(OUTRIDER)
```

### rep dfs

creating dfs of technical replicates - 8 technical reps - 3 batches - muscle/blood

```{r}
test_meta <- fread("test_metadata_DGE_checkpoint.csv")
meta_df <- test_meta[,1:7]
meta_df
```
```{r}
test_count_mat <- fread('final_count_mat.csv')
test_counts <- data.frame(test_count_mat)
test_counts

#check for any dups - if any get index
anyDuplicated(test_counts$Geneid)
test_counts$Geneid[which(duplicated(test_counts$Geneid))]
```
```{r}
#drop dup and impute null counts as 0
test_dup <- test_counts[-14998, ]
test_dups <- data.frame(test_dup, row.names = 1)
test_dups[is.na(test_dups)] <- 0
test_dups
```

###  Collapse replciates and factor meta
```{r}
#factor cols and relevel
meta_df$BATCH<- factor(meta_df$BATCH)

meta_df$COLLECTION_TYPE <- factor(meta_df$COLLECTION_TYPE)

meta_df$MND <- factor(meta_df$MND)

meta_df$AGE <- factor(meta_df$AGE)

meta_df$SEX <- factor(meta_df$SEX)

meta_df$ID <- factor(meta_df$ID)
```
```{r}
#create deseq2 matrix and design 
dds <- DESeqDataSetFromMatrix(countData = test_dups,
                              colData = meta_df,
                              #account for batch effect 
                              design = ~ BATCH + SEX + MND)

#collapse technical replicates
dds_cols <- collapseReplicates(dds, dds$ID, dds$RUN)
dds_cols
```

### Data preprocessing and exploratory analyses 

```{r}
#filter lowly expressed genes 
keep <- rowSums(counts(dds_cols) >= 10) >= 3
dds_cols <- dds_cols[keep,]
dds_cols

#re-level and set MND status ref as control 
dds_cols$MND<- relevel(dds_cols$MND, ref = "Control")
dds_cols$MND
```
## Expression distribution

```{r}
#estimate library sizes 
est_dds <- estimateSizeFactors(dds_cols)

myColors <- hue_pal()(4)

count_s1 <- counts(est_dds, normalized=TRUE)
## Plot the log2-transformed data with a +1 pseudocount
plotDensity(log2(count_s1 + 1), col=rep(myColors, each=3),
            lty=c(1:ncol(count_s1 )), xlab='Log2(normalised count + 1)',
            main='Expression Distribution')
```

## PCA and experiment summary

```{r}

#normalise data and summarise cols for PCA
se <- SummarizedExperiment(log2(counts(est_dds, normalized=TRUE) + 1),
                           colData=colData(est_dds))
#plot PCA 
pca_d1 <- plotPCA(DESeqTransform(se), intgroup=c("COLLECTION_TYPE", "SEX", "MND"), returnData=TRUE)

percentVar <- round(100 * attr(pca_d1, "percentVar"))
pca_plot <- ggplot(pca_d1, aes(PC1, PC2, color=MND, shape=SEX, group=COLLECTION_TYPE)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of PAX Blood vs Muscle Biopsy Samples") + 
  theme(plot.title = element_text(hjust = 0.5)) + stat_ellipse(linetype=2, aes(PC1, PC2, color=COLLECTION_TYPE)) + stat_ellipse(linetype=2, aes(PC1, PC2, color=COLLECTION_TYPE)) 
```



```{r}
pca_plot
```

### check for batch effects - VST

```{r}
vst <- vst(est_dds)

plotPCA(vst, intgroup=c("BATCH", "MND"))

assay(vst) <- limma::removeBatchEffect(assay(vst), vst$BATCH)

plotPCA(vst, intgroup=c("BATCH", "MND"))
```
##Filt Blood & Muscle

```{r}
filt_b <- est_dds[,est_dds$COLLECTION_TYPE == "Pax"]

#filter muscle
filt_mu <- est_dds[,est_dds$COLLECTION_TYPE == "Muscle"]
filt_mu
```
##PCA Blood and Muscle No Batch Effect
```{r}
vst_blood <- vst(filt_b)

plotPCA(vst_blood, intgroup=c( "MND", "SEX"))

assay(vst_blood) <- limma::removeBatchEffect(assay(vst_blood), vst_blood$BATCH)

blood_pca <- plotPCA(vst_blood, intgroup=c( "MND", "SEX"), returnData = T)

percentVar <- round(100 * attr(blood_pca, "percentVar"))
pca_blood <- ggplot(blood_pca, aes(PC1, PC2, color=MND, shape=SEX)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of Blood") + 
  theme(plot.title = element_text(hjust = 0.5)) + theme_bw() 
  #stat_ellipse(
   # data = blood_pca, aes(fill = factor(MND)),
   # geom = "polygon", alpha = 0.1, level = .95, type = "t", linetype = 2, show.legend = F
 # ) 

##PCA muscle 
vst_muscle <- vst(filt_mu)

plotPCA(vst_muscle, intgroup=c( "MND", "SEX"))

assay(vst_muscle) <- limma::removeBatchEffect(assay(vst_muscle), vst_muscle$BATCH)

muscle_pca <- plotPCA(vst_muscle, intgroup=c( "MND", "SEX"), returnData = T)

percentVar <- round(100 * attr(muscle_pca, "percentVar"))
pca_Muscle <- ggplot(muscle_pca, aes(PC1, PC2, color=MND, shape=SEX)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + 
  ggtitle("PCA of Muscle") + 
  theme(plot.title = element_text(hjust = 0.5)) + theme_bw()
#+ 
 # stat_ellipse(
    #data = muscle_pca, aes(fill = factor(MND)),
    #geom = "polygon", alpha = 0.1, level = .95, type = "t", linetype = 2, show.legend = F
 # ) 

#join plot
ggarrange(pca_blood, pca_Muscle, ncol =2, align = "h", common.legend = TRUE, legend="bottom")
```
##ALS genes from literature - Case Vs Control 

```{r}

#a <- filt_ensembl[filt_ensembl$external_gene_name == c('FUS'),]
#a
#UNC13A, SOD1, TARDBP, FUS 
check.vec <- c("ENSG00000120948.19", "ENSG00000130477.16", "ENSG00000142168.15", "ENSG00000089280.19")
check.vec

#transpose and get norm logged counts 
tcounts <- t(log2((counts(est_dds[check.vec, ], normalized=TRUE, replaced=FALSE)+1))) %>%
  merge(colData(est_dds), ., by="row.names") %>%
  gather(gene, expression, (ncol(.)-length(check.vec)+1):ncol(.))

#add gene annotation
t <- tcounts %>% mutate(ensembl_gene_id = gsub("\\..*","", gene))

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
listDatasets(ensembl) %>% filter(grepl("Human",description))

ensembl.con <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", host="https://useast.ensembl.org")
attr <- listAttributes(ensembl.con)
filters <- listFilters(ensembl.con)

t_names <- as.data.frame(getBM(attributes = c('ensembl_gene_id', 'external_gene_name'),
      filters="ensembl_gene_id",
      values = t$ensembl_gene_id,
      mart = ensembl.con))

merged_T <- merge(t, t_names, by='ensembl_gene_id', all='TRUE')
merged_T

#plot and wrap
t_names
p <- ggplot(merged_T, aes(MND, expression, fill=COLLECTION_TYPE)) + 
  geom_boxplot() + 
  facet_wrap(~external_gene_name, scales="free_y") + 
  labs(x="ALS status", 
       y="Expression (log normalised counts)", 
       fill="Tissue")

#t-test with pairwise comparison 
comparison <- list(c("Control", "Case"))
p + stat_compare_means(method = "t.test", label.y = 11.5, label = "p.format") + stat_compare_means(comparisons = comparison) + theme_bw()
```
##DGE Analysis - Whole Dataset


```{r}
deseqdds <- DESeq(est_dds)
``` 

```{r}
dseq_res <- DESeq2::results(deseqdds)
```

```{r}
summary(dseq_res)
```
```{r}
resultsNames(deseqdds)
```
```{r}
res.05 <- DESeq2::results(deseqdds)
table(res.05$padj < 0.05)
```
```{r}
#set dseq results as df
dseq_df <- as.data.frame(dseq_res)

head(dseq_df)
```

```{r}
#use tibble to set rowname to col for edit
df <- tibble::rownames_to_column(dseq_df, "ID")
#need to be trimmed using regex and tidy for ensembl ids 
count_trim <- df %>% mutate(ensembl_gene_id = gsub("\\..*","", ID))
nrow(count_trim)
head(count_trim)

#BIomart db - get annotations 
listMarts()

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
listDatasets(ensembl) %>% filter(grepl("Human",description))

ensembl.con <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", host="https://useast.ensembl.org")
attr <- listAttributes(ensembl.con)
filters <- listFilters(ensembl.con)

#id, gene type, gene name
gene_df <- as.data.frame(getBM(attributes = c('ensembl_gene_id',"gene_biotype", 'external_gene_name'),
      filters="ensembl_gene_id",
      values = count_trim$ensembl_gene_id,
      mart = ensembl.con))

#merge dfs via ensemble ids 
ensembl_DEGs <- merge(count_trim, gene_df, by='ensembl_gene_id', all='TRUE')
#filt nans
filt_ensembl <- drop_na(ensembl_DEGs)
filt_ensembl <- data.frame(filt_ensembl, row.names = 2)
#filt_ensembl <- filt_ensembl %>% distinct(external_gene_name, .keep_all = TRUE)
head(filt_ensembl)

```


## Upregulated and downregulated genes

```{r}
#upregulated genes at LFC > 1 threshold and sig by p adj
upreg <- filt_ensembl %>% filter(log2FoldChange > 1 & padj < 0.05)
#no upreg genes
nrow(upreg)
upreg
downreg <- filt_ensembl %>% filter(log2FoldChange < -1 & padj < 0.05)
nrow(downreg)

```
```{r}
#dispersion plot
plotDispEsts(deseqdds, main='Dispersion Plot')
```
```{r}
DESeq2::plotMA(dseq_res, ylim=c(-2,2))
```

##Hypothesis weighting 

```{r}
#suppressPackageStartupMessages(library(IHW))

#res.ihw <- DESeq2::results(deseqdds, filterFun = ihw)
#summary(res.ihw)
```
```{r}
#ihw.obj <- metadata(res.ihw)$ihwResult
#plot(ihw.obj)
#plot(ihw.obj, what = "decisionboundary") 
```

##Whole data - Top 50 DEGs

```{r}
#subset to get only protein coding genes
pc_genes <- subset(filt_ensembl, gene_biotype=='protein_coding')

#get top 50 protein coding DEGs
top_50_genes <- head(pc_genes[order(pc_genes$padj),], 50)
#set as df
top_50_genes <- as.data.frame(top_50_genes)
top_50_genes
```
##Volcano Plot of PCGs

```{r Volcano1, fig.height = 8, fig.width = 6.5, fig.align = "center"}
lab_italics <- paste0("italic('", lab = pc_genes$external_gene_name, "')")
  selectLab_italics = paste0(
    "italic('",
    c('COL19A1', "SAA1","SAA2","CDKN1A", "FGFBP2"),
    "')")

d1 <- EnhancedVolcano(pc_genes,
    lab = lab_italics,
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = selectLab_italics,
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 1e-4,
    title = 'Volcano Plot',
    FCcutoff = 1.0,
    pointSize = 3.0,
    labSize = 6.0,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    parseLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'bottom',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')


d1
```

```{r}
#normalise counts for matrix 
mat <- counts(deseqdds, normalized = T)[rownames(top_50_genes),]
#get zscores for new matrix scale by 1 when transposing 
mat_z <- t(apply(mat,1,scale))
#set col names for new df
colnames(mat_z) <-rownames(dds_cols@metadata)
rownames(mat_z) <- top_50_genes$external_gene_name
```

## DEG Heatmap

```{r}

ann <- data.frame(Sex = dds_cols$SEX,
                  ALS= dds_cols$MND, 
                  Tissue = dds_cols$COLLECTION_TYPE, stringsAsFactors = F)

colours <- list(
    Sex = c('Male' = 'blue', 'Female' = 'green'),
    ALS = c("Case" = "yellow","Control" = "orange"),
    Tissue = c("Pax" = "purple", "Muscle" = "pink"))
colAnn <- HeatmapAnnotation(
    df = ann,
    which = "col", # 'col' (samples) or 'row' (gene) annotation?
    na_col = 'white', # default colour for any NA values in the annotation data-frame, 'ann'
    col = colours,
    annotation_height = 0.6,
    annotation_width = unit(1, 'cm'),
    gap = unit(1, 'mm'),
    annotation_legend_param = list(
      Sex = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'Sex',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface='bold'),
        labels_gp = gpar(fontsize = 8)),
      ALS = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'ALS',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface = "bold"),
        labels_gp = gpar(fontsize = 8)),
      Tissue = list(
        nrow = 1, # number of rows across which the legend will be arranged
        title = 'Tissue',
        title_position = 'topcenter',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface = "bold"),
        labels_gp = gpar(fontsize = 8))
      ))

#make heatmap for clustering 
DGE_heatmap <- Heatmap(mat_z, 
                       cluster_rows = T, 
                       cluster_columns = T, 
                       column_labels = colnames(mat_z),
                       name = "z-score",
                       column_title = "Samples",
                       row_title = "Genes", 
                       show_row_names = T,
                       show_column_names = F,
                       top_annotation = colAnn)

DGE_heatmap
```
##Check DEG SAA1, SAA2, SAA2-SAA4
```{r}
##Investigate SAA1, SAA2, SAA2-SAA4 
#a <- filt_ensembl[filt_ensembl$external_gene_name == c('MYH3'),]
#a
#
check.vec_SAA <- c("ENSG00000173432.13", "ENSG00000134339.9", "ENSG00000255071.3")
check.vec_SAA

#transpose and get norm logged counts 
tcounts_saa <- t(log2((counts(est_dds[check.vec_SAA, ], normalized=TRUE, replaced=FALSE)+1))) %>%
  merge(colData(est_dds), ., by="row.names") %>%
  gather(gene, expression, (ncol(.)-length(check.vec_SAA)+1):ncol(.))

tcounts_saa

#add gene annotation
t_saa <- tcounts_saa %>% mutate(ensembl_gene_id = gsub("\\..*","", gene))
t_saa

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
listDatasets(ensembl) %>% filter(grepl("Human",description))

ensembl.con <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", host="https://useast.ensembl.org")
attr <- listAttributes(ensembl.con)
filters <- listFilters(ensembl.con)

t_names_saa <- as.data.frame(getBM(attributes = c('ensembl_gene_id', 'external_gene_name'),
      filters="ensembl_gene_id",
      values = t_saa$ensembl_gene_id,
      mart = ensembl.con))
t_names_saa

merged_T_saa <- merge(t_saa, t_names_saa, by='ensembl_gene_id', all='TRUE')
merged_T_saa

#plot and wrap
p_saa <- ggplot(merged_T_saa, aes(MND, expression, fill=SEX)) + 
  geom_boxplot() + 
  facet_wrap(~external_gene_name, scales="free_y") + 
  labs(x="ALS status", 
       y="Expression (log normalised counts)", 
       fill="Sex")


#t-test with pairwise comparison 
comparison_saa <- list(c("Control", "Case"))
p_saa + stat_compare_means(method = "t.test", label.y = 11.5, label = "p.format") + stat_compare_means(comparisons = comparison_saa) + theme_bw()


```



##GO - DEGs

```{r}
GO_DEGs <- top_50_genes$ensembl_gene_id
GO <- enrichGO(gene = GO_DEGs, OrgDb = "org.Hs.eg.db", key = "ENSEMBL", ont = "ALL")
GO %>% as.data.frame()
GO
dotplot(GO, title=" Differentially Expressed Genes") + facet_grid(ONTOLOGY~., scales="free")
```
## GO - Upregulated Genes
```{r}
pc_upreg <- upreg %>% filter(gene_biotype == 'protein_coding')
GO_upreg <- pc_upreg$ensembl_gene_id
GO_d <- enrichGO(gene = GO_upreg, OrgDb = "org.Hs.eg.db", key = "ENSEMBL", ont = "ALL")
GO_d %>% as.data.frame()
dotplot(GO_d, title="Upregulated Differentially Expressed Genes")
```

## GO - Downregulated Genes
```{r}
pc_downreg <- downreg %>% filter(gene_biotype == 'protein_coding')
GO_downreg <- pc_downreg$ensembl_gene_id
GO_d <- enrichGO(gene = GO_downreg, OrgDb = "org.Hs.eg.db", key = "ENSEMBL", ont = "ALL")
GO_d %>% as.data.frame()
dotplot(GO_d, title="Downregulated Differentially Expressed Genes")
```

##GSEA
```{r}
#get top 50 protein coding DEGs
gsea_genes<- head(pc_genes[order(pc_genes$padj),], 200)
gsea_genes
```
```{r}
organism = "org.Hs.eg.db"
# we want the log2 fold change 
original_gene_list <- gsea_genes$log2FoldChange

# name the vector
names(original_gene_list) <- gsea_genes$ensembl_gene_id

# omit any NA values 
gene_list<-na.omit(original_gene_list)

# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)

gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "none")
```

```{r}
require(DOSE)
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
```
```{r}
library(enrichplot)
gse$Description
gseaplot2(gse, title = gse$Description[36], geneSetID = 36)
```

```{r}
library(ggupset)
edox <- setReadable(gse, 'org.Hs.eg.db', 'auto')
edox
p1 <- cnetplot(edox, foldChange=gene_list)
p1
```
```{r}
upsetplot(gse)
```



```{r}
ridgeplot(gse) + labs(x = "enrichment distribution")
```

###KEGG

```{r}
# Convert gene IDs for gseKEGG function
# We will lose some genes here because not all IDs will be converted
ids<-bitr(names(original_gene_list), fromType = "ENSEMBL", toType = "ENTREZID", OrgDb=organism)
 # remove duplicate IDS (here I use "ENSEMBL", but it should be whatever was selected as keyType)
dedup_ids = ids[!duplicated(ids[c("ENSEMBL")]),]

# Create a new dataframe df2 which has only the genes which were successfully mapped using the bitr function above
df2 = gsea_genes[gsea_genes$ensembl_gene_id %in% dedup_ids$ENSEMBL,]

# Create a new column in df2 with the corresponding ENTREZ IDs
df2$Y = dedup_ids$ENTREZID

# Create a vector of the gene unuiverse
kegg_gene_list <- df2$log2FoldChange

# Name vector with ENTREZ ids
names(kegg_gene_list) <- df2$Y

# omit any NA values 
kegg_gene_list<-na.omit(kegg_gene_list)

# sort the list in decreasing order (required for clusterProfiler)
kegg_gene_list = sort(kegg_gene_list, decreasing = TRUE)
kegg_gene_list
```
```{r}
kegg_organism = "hsa"
kk2 <- gseKEGG(geneList     = kegg_gene_list,
               organism     = kegg_organism,
               nPerm        = 10000,
               minGSSize    = 3,
               maxGSSize    = 800,
               pvalueCutoff = 0.05,
               pAdjustMethod = "BH")

bitr_kegg(names(kegg_gene_list), fromType = "kegg", toType = "Path", organism = "hsa")

kk3 <- enrichKEGG(gene = names(kegg_gene_list), org = 'hsa')
kk3
```



